DOpElib: Differential Equations\\ and Optimization Environment

### 2013-12-09 Examples/PDE/StatPDE/Example5/content.tex

\textbf{General problem description}

\vspace{0.2cm}

Similar to the previous example, we consider the following benchmark problem from plasticity theory:
\begin{equation}
   (\Pi(\sigma(u)),\varepsilon(\varphi)) = (g,\varphi)_{\Gamma_N}.
\end{equation}
Here $\tilde{\Omega}$ is again the quadratic domain with a circular hole around the center cut out. Again, we restrict our actual computational domain $\Omega$ to the upper left quarter of $\tilde{\Omega}$ for reasons of symmetry.\\
We use the symmetric strain tensor $\varepsilon(v) := \frac{1}{2}(\nabla v + \nabla v^T)$, and the symmetric stress tensor $\sigma$ is defined as
\begin{equation*}
   \sigma(v) := 2\mu \varepsilon(v)^D + \kappa \cdot tr (\varepsilon(v)) I
\end{equation*}
where $\tau^D$ is the deviatoric part of a tensor $\tau$, in two dimensions defined as
\begin{equation*}
   \tau^D := \tau - \frac{1}{2} tr(\tau) I.
\end{equation*}
The main difference with respect to the elastic case is the projection operator $\Pi$ in equation $(1)$. It is defined as follows:
\begin{equation*}
\Pi(\tau) = \left\{
            \begin{array}{lr}
            \tau & |\tau^D| \leq \sigma_0\\
            \sigma_0 |\tau^D|^{-1} \tau^D + \frac{1}{2} tr(\tau) I & |\tau^D| > \sigma_0\\ 
            \end{array}
            \right. 
\end{equation*}
In our computations, we choose $\sigma_0 = \sqrt{\frac{2}{3}}\cdot 450$, and the above parameters $\mu$ and $\kappa$ as $\mu = 80193.800283$ resp. $\kappa = 190937.589172$. The corner points of our computational domain are the same as before, and the boundary conditions are not altered, either.\\
The goal of our computations is to detect a subdomain in $\Omega$ where plastic behaviour occurs (compare \textit{E. Stein (editor), Error-controlled Adaptive Finite Elements in Solid Mechanics, Wiley (2003), pp. 386 - 389}). This subdomain depends on the righthand side $g$ in equation $(1)$ which we write as $g = \lambda \cdot p$ with $p = 100$ and $\lambda \in [1.5;4.5]$.\\

\vspace{0.2cm}

\textbf{Program description}

\vspace{0.2cm}

The code of the current example is nearly identical to the code of the previous one. The only difference worth mentioning is the change of the equations which leads to different implementations of the \texttt{CellEquation, CellMatrix} and \texttt{BoundaryEquations} functions in \textit{localpde.h}.\\
Furthermore, the elasticity equations solved in the last example are linear, whereas the plasticity equations are nonlinear; this difference is evident also from the output (here, we need several Newton steps until convergence).\\
The functionals that appear in the output yield additional information and are not required in the above problem setting. The subdomain with plastic behaviour we want to detect can be visualized from the \textit{.vtk} files written to the \texttt{Results/Mesh} subfolders.

>>>>

\subsubsection{General problem description}
This example shows the use of the adaptive grid refinement and error estimation by the \textit{DWR method} (For a description of the method, see \cite{BR03}.) applied to the laplace equation 
\begin{align*}
-\Delta u &= f && \text{ in } \Omega
\end{align*}
with the analytical solution
\begin{align*}
u = \sin\left(\frac{\pi}{x^2 + y^2}\right),
\end{align*}
 the corresponding right hand side $f = -\Delta u$ and appropriate Dirichlet Conditions on $\partial \Omega$, where the domain is given by
\begin{align*}
\Omega = [-2,2]^2\setminus \overline{B}_{0.5}(0).
\end{align*}
We want to estimate the error in the following functional of interest
$$\fullfunction{J}{H^1(\Omega)}{\mathbb{R}}{u}{\frac 1 {\abs{\Gamma}} \int_\Gamma u \,dx}$$
where
$\Gamma = \left\{(x,y)\in \mathbb R ^2\,\big | \,x=0,-2<y<0.5 \right \}$.

For this setting, we have the error representation
\begin{align}\label{dwr.error_identity}
J(e)=\sum_{K\in \mathbb T _h} \left\{(R_h, z-\psi_h)_K + (r_h, z-\psi_h)_{\partial_K}\right\}
\end{align}
with the error $e=u-u_h$, the Triangulation $\mathbb T_h$, the dual solution $z$, arbitraty function $\phi_h \in V_h$ (the ansatz space) and the element- and edge-residuals: 
\begin{align}\label{dwr.element_residual}
\eval[1]{R_h}_{K} = f+\Delta u_h
\end{align}
resp. 
\begin{align}\label{dwr.edge_residual}
\eval[1]{r_h}_{\Sigma} =
\begin{cases}
\frac 12 [\partial_n u_h] ,&\text{if } \Sigma \subset \partial_K \setminus \partial \Omega,\\
0,& \text{if } \Sigma \subset\partial \Omega.
\end{cases}
\end{align}

It holds $J(u) \approx 0.441956231972232$.
\subsubsection{Program description}
In this section we want to focus on what you have to do if you want to enhance your existing code to use the \textit{DWR method}.

First, additionally to all the things one has to do when just solving the equation, we have to include the file
\begin{verbatim}
higher_order_dwrc.h
\end{verbatim}

As we approximate the so called 'weights' $z-\phi_h$ in the error representation by a patchwise higher order interpolation of $z_h$ (the computed dual solution), we have to enforce patch-wise refinement of the grid by giving the flag 
\begin{verbatim}
Triangulation<2>::MeshSmoothing::patch_level_1
\end{verbatim}
to the triangulation.

To be able to solve the adjoint equation for the error estimation one needs to implement some methods regarding the equation as well es the functional of interest:
\begin{itemize}
\item In {\verb pdeinterface.h }
\begin{itemize}
\item {\verb ElementEquation_U }: Weak form of the adjoint equation.
\item {\verb ElementMatrix_T }: The FE matrix for the adjoint problem.
\item {\verb FaceEquation_U }: This one is needed in this case here because we have a functional of interest that lives on faces. 
\end{itemize}
\item {\verb functionalinterface.h }
\begin{itemize}
\item {\verb FaceValue_U }: This is the right hand side of the adjoint equation.
\end{itemize}
\end{itemize}
During the  evaluation of \eqref{dwr.error_identity}, the following methods are needed
\begin{itemize}
\item {\verb StrongElementResidual }: The element residual, see \eqref{dwr.element_residual}.
\item {\verb StrongFaceResidual }: The terms in \eqref{dwr.error_identity} that lies in the interiour (i.e. the jumps).
\item {\verb StrongBoundaryResidual }:The terms in \eqref{dwr.error_identity} that lies on the boundary (There are none in this case).
\end{itemize}

Note that in the above three functions we always apply the method \texttt{ResidualModifier} both to the 
residual as well as to the jumps on the faces. This is done to assert that we can apply both a 
DWR-error estimator where the residual should be multiplied with the computed weights (then this function 
does not do anything) as well as Residual Type error estimator for the $L^2$ or $H^1$ norm where we need to calculate element wise norms of the residual and the jumps. Then this function calculates the appropriate 
local terms, e.g., the square of the residual scaled with appropriate powers of the local mesh size.

After this, we tell the problem which functional we want to use for the error estimation, this is done via
\begin{verbatim}
P.SetFunctionalForErrorEstimation(LFF.GetName())
\end{verbatim}
where \texttt{P} is of type \texttt{PDEProblemContainer} and \texttt{LFF} is the desired functional of interest.

The next thing we need is an object of the type
\begin{verbatim}
HigherOrderDWRContainer
\end{verbatim}
This container takes care of the computation of the weights.

To build this, we need the following:
\begin{itemize}
\item {\verb DOFH_higher_order }: With some higher order Finite Elemnts and the already defined triangulation, we build this \texttt{SpaceTimeHandler}. This is needed because  we want to use the patch-wise higher order interpolation of the weights.
\item {\verb idc_high }: A \texttt{IntegratorDataContainer} in which we put some (face)quadrature formulas for the evaluation of the error Identity.
\item A string which indicates how we want to store the weight-vectors (here: { \verb "fullmem" }).
\item \texttt{pr}: The \texttt{ParameterReader} which we have already defined.
\item A enum of type \texttt{EETerms} that tells the container, which error terms we want to compute (primal error indicators vs. dual error indicators, see \cite{BR03}).  
\end{itemize}
The last preparation step is now to initialize the \texttt{DWRDataContainer} with the problem in use:
\begin{verbatim}
  solver.InitializeHigherOrderDWRC(dwrc);
\end{verbatim}

Succeding the solution of the state equation 
\begin{verbatim}
solver.ComputeReducedFunctionals(); 
\end{verbatim}
we compute the error indicators by calling 
\begin{verbatim}
solver.ComputeRefinementIndicators(dwrc);
\end{verbatim}
We can now get the error indicators (with signs!) out of \texttt{dwrc} by 
\begin{verbatim}
dwrc.GetErrorIndicators();
\end{verbatim}
With these indicators, we are now able to refine our grid adaptively (there are several mesh adaption strategies implemented, like 'RefineOptimized', 'RefineFixedNumber' or 'RefineFixedFraction')
\begin{verbatim}
DOFH.RefineSpace(RefineOptimized(optimized, error_ind));
\end{verbatim}
Note, that one has to take the norm of each entry in the vector of the error indicators before feeding them into the \texttt{RefineSpace} method.

### 2013-12-09 Examples/PDE/StatPDE/Example6/content.tex
Further work on the examples

todo. short: this example shows how to use the constraintsmaker class by enforcing periodic boundary conditions in 2d.

>>>>

\subsubsection{General problem description}
In this problem we solve the simple vector valued laplace equation in 3d
on the unit square $\Omega=[0,1]^3$, i.e. in strong formulation we look for $u=(u_1, u_2, u_3)$ s.t.
\begin{align*}
-\Delta u =& f &&\text{in }\Omega.
\end{align*}
We set zero dirichlet values on $\partial \Gamma$ and choose $f=(1,1,1)$.
\subsubsection{Program description}
The PDE is discretized with Q3-elements on a series of locally refined grids (we use the \texttt{KellyErrorEstimator}). The algebraic equations are solved with  different iterative linear solvers acting on different vector and matrix-structures (i.e. we use \texttt{dealii::BlockVector} and \texttt{dealii::Vector} plus the appropriate matrix classes).

To switch the linear solver is pretty easy since the newton solver has a template for the linear solver. Thus changing this template is all that is required.

To change the structure of the vectors and matrices involves also only the change of some template parameters. Our example programs are mostly build such that only a change of some typedefs is required, i.e. one has to interchange the lines
\begin{verbatim}
	typedef SparseMatrix<double> MATRIX;
	typedef SparsityPattern SPARSITYPATTERN;
	typedef Vector<double> VECTOR;
\end{verbatim}
with
\begin{verbatim}
	typedef BlockSparseMatrix<double> MATRIX;
	typedef BlockSparsityPattern SPARSITYPATTERN;
	typedef BlockVector<double> VECTOR;
\end{verbatim}
to switch between the block and non-block structures.

After solving the equation, we want to apply local mesh refinement. So first we extract with the help of the \texttt{SolutionExatractr}-class the vector \texttt{solution} representing the finite element solution
\begin{verbatim}
  SolutionExtractor<SSolver1, VECTORBLOCK> a1(solver1);
 	const StateVector<VECTORBLOCK> &gu1 = a1.GetU();
 	solution = gu1.GetSpacialVector();
\end{verbatim}
With this vector we estiamte the error via \texttt{KellyErrorEstimator} and get a vector holding the estimated error per element. After choosing a refinement criterion (see \textit{refinementcontainer.h}, we opt here for refining the top 20\% of the elements), we give the SpaceTimeHandler an object of type \texttt{RefinementContainer} which holds all the information needed for the local mesh refinement. This is done via the \texttt{RefineSpace} method.
\begin{verbatim}
        DOFH1.RefineSpace(
               RefineFixedNumber(estimated_error_per_element, 0.2, 0.0));
\end{verbatim}
This method transfers our solution onto the new mesh. The transferred solution is then taken as the starting guess of the newton method in the next solution cycle. This is especially helpful for nonlinear problems.

### 2013-12-09 Examples/PDE/StatPDE/Example7/content.tex

\subsubsection{General problem description}
This is the simple vector valued laplace equation on the unitsquare with righthandside $(1,1)$ and zero dirichlet values. The classical example of a PDE.

>>>>

\subsubsection{General problem description}

In this example we consider the following benchmark problem from elasticity theory: \index{stationary PDE} \index{stationary PDE!elasticity equations}
\begin{equation}
   (\sigma(u),\varepsilon(\varphi)) = (g,\varphi)_{\Gamma_N}.
\end{equation}
Here $\tilde{\Omega}$ is a quadratic domain with side length $200$ mm, where a circular hole with radius $10$ mm around the center is cut out. Using symmetries of the domain, we restrict our actual computational domain $\Omega$ to the upper left quarter of $\tilde{\Omega}$.\\
In the above equation, $\varepsilon(v) := \frac{1}{2}(\nabla v + \nabla v^T)$ is the symmetric strain tensor, and \index{strain tensor}
\begin{equation*}
   \sigma(v) := 2\mu \varepsilon(v)^D + \kappa \cdot tr (\varepsilon(v)) I
\end{equation*}
denotes the symmetric stress tensor. Here $\tau^D$ is the deviatoric part of a tensor $\tau$, in two dimensions defined as \index{deviator}
\begin{equation*}
   \tau^D := \tau - \frac{1}{2} tr(\tau) I,
\end{equation*}
and the parameters $\mu$ and $\kappa$ are chosen as $\mu = 80193.800283$ resp. $\kappa = 190937.589172$.\\
The corner points of our computational domain are in anticlockwise order: $(0,0), (90,0)$, $(100,10), (100,100)$ and $(0,100)$. We prescribe homogeneous Dirichlet boundary conditions in $y$-direction between $(0,0)$ and $(90,0)$ (lower boundary part), homogeneous Dirichlet boundary conditions in $x$-direction between $(100,10)$ and $(100,100)$ (right boundary part), and we interpret the righthand side of equation $(1)$ with $g = 450$ as a boundary condition between $(0,100)$ and $(100,100)$ (upper boundary part).\\
The goal of our computations is to match the following functional reference values taken from \textit{E. Stein (editor), Error-controlled Adaptive Finite Elements in Solid Mechanics, Wiley (2003), pp. 386 - 387}:
\begin{table}[h]
\centering
\begin{tabular}{lccc}    
\hline
 Functional & $u_1$ at $(90,0)$ & $\sigma_{22}$ at $(90,0)$ & $u_2$ at $(100,100)$\\
\hline 
 Reference value & 0.021290 & 1388.732343 & 0.20951\\
\hline
\end{tabular}
\end{table}
\begin{table}[h]
\centering
\begin{tabular}{lcc}    
\hline
 Functional & $u_1$ at $(0,100)$ & $\int_{(100,100)}^{(0,100)} u_2$\\
\hline 
 Reference value & 0.076758 & 20.40344\\
\hline
\end{tabular}
\end{table}

\subsubsection{Program description}

From the previous examples we know how to read a grid from an \textit{.inp} file. The grid of our current example comes from the above mentioned benchmark problem. The parameter file (\textit{dope.prm}) contains the same variables as in the first two examples.\\
Apart from different point values of derivatives of the solution, we want to evaluate an integral over part of the boundary. \index{functional} \index{functional!boundary integral} This is newly implemented in \textit{functionals.h}.\\
The (linear) elasticity equation is much simpler than the FSI problems we treated before, which can be seen in \textit{localpde.h} where the \texttt{ElementEquation} and \texttt{ElementMatrix} are implemented. In principle, everything is clear from the preceding examples.\\
The \textit{main.cc} file does not contain any innovation, either. Here, we refine the grid globally instead of using an error estimator for local refinement.\\
The output of the program reflects again the linearity of the problem (only one Newton step is needed for solution).\\

### 2013-12-09 Examples/PDE/StatPDE/Example8/content.tex

The code of the current example is nearly identical to the code of the previous one. The only difference worth mentioning is the change of the equations which leads to different implementations of the \texttt{CellEquation, CellMatrix} and \texttt{BoundaryEquations} functions in \textit{localpde.h}.\\
Furthermore, the elasticity equations solved in the last example are linear, whereas the plasticity equations are nonlinear; this difference is evident also from the output (here, we need several Newton steps until convergence).\\
The functionals that appear in the output yield additional information and are not required in the above problem setting. The subdomain with plastic behaviour we want to detect can be visualized from the \textit{.vtk} files written to the \texttt{Results/Mesh} subfolders.

>>>>

The code of the current example is nearly identical to the code of the previous one. The only difference worth mentioning is the change of the equations which leads to different implementations of the \texttt{ElementEquation, ElementMatrix} and \texttt{BoundaryEquations} functions in \textit{localpde.h}.\\
Furthermore, the elasticity equations solved in the last example are linear, whereas the plasticity equations are nonlinear; this difference is evident also from the output (here, we need several Newton steps until convergence).\\
The functionals that appear in the output yield additional information and are not required in the above problem setting. The subdomain with plastic behaviour we want to detect can be visualized from the \textit{.vtk} files written to the \texttt{Results/Mesh} subfolders.
